---
title: "Making multivariate statistics reproducible"
output: html_notebook
thanks: "Replication files are available on the author's Github account."
author:
- name: Felix Taschbach
  affiliation: Maastricht University
---
## Load Packages
We begin by installing and loading any required packages. The following code first installs the package pacman, if necessary, and then uses pacman to install, if required, and load the packages needed to run this script.
```{r}
# Clear all variables 
rm(list=ls()) 

# install pacman if necessary
if (!require("pacman")) suppressPackageStartupMessages(install.packages("pacman"))
# install required packages
pacman::p_load("ggplot2", "dplyr","pls","readr","caTools","caret","gplots","car","rcdk","iterators","lattice","latticeExtra")
```
## Importing the data.
To import the data I am first setting the local path to the directory that the data is stored in.
```{r}
DATA.DIR <- "/Users/felix/Google Drive/Uni/Systems Biology/Year 2/Scientific Programming/scientificProgramming2/data"
setwd(DATA.DIR) # set the working directory
```

Afterwards, I import the activity scores. To properly import the activity scores matrix, 2 import function calls are needed: one to import the header and one to import the data.
```{r}
# import the header and the activity scores
headers <- read.csv('aid624202.csv', header=FALSE, nrows=1, as.is=TRUE, row=1)
data.raw <- read.csv('aid624202.csv',skip=6,header=FALSE,row=1)
colnames(data.raw) <- headers                                       # change the names of the data to the correct ones
data.raw <- data.raw[,c("PUBCHEM_SID","PUBCHEM_ACTIVITY_SCORE")]     # select SID and the activity score from the data
colnames(data.raw) <- c("SID","Score")                               # change column names
```
